---
title: "Joint Representation of Spliced and Unspliced RNA: Analysis Driver"
author: "Arya Jadhav and Zachary J. DeBruine"
output:
  html_document:
    toc: true
    toc_depth: 2
---

# Overview

This notebook provides a reusable analysis workflow for the spliced,
unspliced, added, and concatenated single-cell RNA representations used in
the manuscript.

To analyze a different dataset, update the dataset configuration below.
The analysis functions themselves are defined in the `code/` directory.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(Matrix)
library(Seurat)
library(singlet)
library(dplyr)
library(tibble)
library(ggplot2)
library(cowplot)
library(fgsea)
library(msigdbr)

source("../code/01_preprocess.R")
source("../code/02_cross_validation.R")
source("../code/03_fit_models.R")
source("../code/04_factor_matching.R")
source("../code/05_gsea.R")
source("../code/06_pca.R")
source("../code/07_figures.R")
source("../code/08_factor_similarity.R")
source("../code/09_modality_loadings.R")
source("../code/10_umap.R")
```

# Dataset configuration

Change these values to run the workflow on another dataset.

```{r configuration}
dataset_name <- "heart"

data_dir <- "../data/heart_10k_v3_cr_extracted"

dataset_prefix <- "heart_10k_v3_cr"

species <- "Mus musculus"

output_dir <- file.path(
  "../results",
  dataset_name
)

dir.create(
  output_dir,
  recursive = TRUE,
  showWarnings = FALSE
)
```

# Preprocessing

Read the spliced and unspliced matrices, apply cell-level quality control,
construct the added and concatenated representations, and normalize the
resulting matrices.

```{r preprocessing}
dataset <- prepare_dataset(
  data_dir = data_dir,
  dataset_prefix = dataset_prefix
)

dataset
```

# Cross-validation

NMF rank is evaluated over ranks 2 through 80.

The `get_best_rank()` result is used as the operational rank for fitting
the downstream NMF models in this workflow. These operational ranks are
distinct from the manually selected revealable-rank values summarized
across replicate cross-validation runs in the manuscript tables.

```{r cross-validation}
set.seed(43)

cv_results <- list(

  spliced = cross_val(
    dataset$concat[
      dataset$spliced_indices,
      ,
      drop = FALSE
    ],
    ranks = 2:80,
    n_replicates = 3
  ),

  unspliced = cross_val(
    dataset$concat[
      dataset$unspliced_indices,
      ,
      drop = FALSE
    ],
    ranks = 2:80,
    n_replicates = 3
  ),

  added = cross_val(
    dataset$added,
    ranks = 2:80,
    n_replicates = 3
  ),

  concat = cross_val(
    dataset$concat,
    ranks = 2:80,
    n_replicates = 3
  )
)
```

# Operational NMF ranks

```{r operational-ranks}
ranks <- c(

  spliced = get_best_rank(
    cv_results$spliced
  ),

  unspliced = get_best_rank(
    cv_results$unspliced
  ),

  added = get_best_rank(
    cv_results$added
  ),

  concat = get_best_rank(
    cv_results$concat
  )
)

ranks
```

# Fit final NMF models

```{r fit-models}
models <- fit_nmf_models(
  dataset = dataset,
  ranks = ranks,
  seed = 123
)
```

# Cross-validation visualization

The plotting function uses the test-error trajectories generated during
cross-validation.

```{r cv-figure}
cv_plot_data <- bind_rows(

  cv_results$spliced %>%
    mutate(
      group = "Spliced Only"
    ),

  cv_results$unspliced %>%
    mutate(
      group = "Unspliced Only"
    ),

  cv_results$added %>%
    mutate(
      group = "Added"
    ),

  cv_results$concat %>%
    mutate(
      group = "Concatenated"
    ) 
) %>%
  group_by(
    group,
    k,
    iter
  ) %>%
  summarise(
    test_err = mean(test_error),
    .groups = "drop"
  )

cv_figure <- plot_cv(
  cv_plot_data
)

cv_figure
```

# PCA scree analysis

```{r pca}
pca_results <- bind_rows(

  get_pca_scree(
    dataset$concat[
      dataset$spliced_indices,
      ,
      drop = FALSE
    ],
    label = "Spliced"
  ),

  get_pca_scree(
    dataset$concat[
      dataset$unspliced_indices,
      ,
      drop = FALSE
    ],
    label = "Unspliced"
  ),

  get_pca_scree(
    dataset$added,
    label = "Added"
  ),

  get_pca_scree(
    dataset$concat,
    label = "Concatenated"
  )
)

pca_figure <- plot_pca_scree(
  pca_results
)

pca_figure
```

# Factor similarity and matching

Compare NMF factors across the four representations using cosine
similarity and optimal bipartite matching.

```{r factor-similarity}
factor_similarity <- factor_similarity_analysis(
  models
)

factor_similarity$figure
```

# Gene-set collection

Construct the MSigDB C5 gene-set collection appropriate for the current
dataset.

```{r gene-sets}
gene_sets <- msigdbr(
  species = species,
  collection = "C5"
) %>%
  dplyr::select(
    gs_name,
    gene_symbol
  ) %>%
  dplyr::distinct() %>%
  dplyr::group_by(
    gs_name
  ) %>%
  dplyr::summarise(
    genes = list(gene_symbol),
    .groups = "drop"
  ) %>%
  tibble::deframe()
```

# Modality loading balance and GSEA

For each concatenated NMF factor, calculate the relative contribution of
spliced and unspliced feature weights and compare enrichment obtained
from the two modality-specific loading vectors.

```{r modality-gsea}
modality_gsea <- plot_modality_balance_gsea(
  concat_nmf = models$concat,
  gene_sets = gene_sets
)

modality_gsea$figure
```

# UMAP comparison

Generate separate UMAP embeddings from the NMF cell-factor matrices for
the four representations. Cluster labels are defined using the Added
representation and transferred to all four embeddings for comparison.

```{r umap}
umap_results <- compare_nmf_umaps(
  dataset = dataset,
  models = models
)

umap_results$figure
```

# Manuscript-specific analyses

Some manuscript figures use selected datasets or selected NMF factors and
are therefore not general steps in the analysis pipeline.

## PBMC selected-factor loading comparison

The factors below correspond to the selected concatenated PBMC factors
shown in Figure 4.

```{r pbmc-selected-factors}
if (tolower(dataset_name) == "pbmc") {

  selected_loadings <- plot_selected_modality_loadings(
    concat_nmf = models$concat,
    spliced_dominant = c(34, 35),
    unspliced_dominant = c(23, 28),
    n_label = 7
  )

  print(
    selected_loadings$figure
  )
}
```

# Save analysis results

```{r save-results}
saveRDS(
  ranks,
  file.path(
    output_dir,
    "operational_ranks.rds"
  )
)

saveRDS(
  cv_results,
  file.path(
    output_dir,
    "cross_validation.rds"
  )
)

saveRDS(
  models,
  file.path(
    output_dir,
    "nmf_models.rds"
  )
)

saveRDS(
  pca_results,
  file.path(
    output_dir,
    "pca_results.rds"
  )
)

saveRDS(
  factor_similarity$matching_results,
  file.path(
    output_dir,
    "factor_matching_results.rds"
  )
)

saveRDS(
  modality_gsea$modality_scores,
  file.path(
    output_dir,
    "modality_loading_scores.rds"
  )
)

saveRDS(
  modality_gsea$gsea_results,
  file.path(
    output_dir,
    "concat_gsea_results.rds"
  )
)

saveRDS(
  umap_results$plot_data,
  file.path(
    output_dir,
    "umap_results.rds"
  )
)
```

# Save figures

```{r save-figures}
ggsave(
  file.path(
    output_dir,
    "cross_validation.pdf"
  ),
  cv_figure,
  width = 12,
  height = 6.5
)

ggsave(
  file.path(
    output_dir,
    "pca_scree.pdf"
  ),
  pca_figure,
  width = 8,
  height = 6
)

ggsave(
  file.path(
    output_dir,
    "factor_similarity.pdf"
  ),
  factor_similarity$figure,
  width = 12,
  height = 6
)

ggsave(
  file.path(
    output_dir,
    "modality_balance_gsea.pdf"
  ),
  modality_gsea$figure,
  width = 11,
  height = 5
)

ggsave(
  file.path(
    output_dir,
    "umap_comparison.pdf"
  ),
  umap_results$figure,
  width = 8,
  height = 7
)

if (
  tolower(dataset_name) == "pbmc" &&
  exists("selected_loadings")
) {

  ggsave(
    file.path(
      output_dir,
      "selected_factor_loadings.pdf"
    ),
    selected_loadings$figure,
    width = 7,
    height = 7
  )
}
```

# Session information

```{r session-info}
sessionInfo()
```
